---
title: "Day1_introduciton"
author: "Marina e."
date: "9/14/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: yeti
    highlight: haddock
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading the data and libraries
```{r}
library(tidyverse)
```

```{r}
blp_df <- read_csv("~/github/rsdwr/data/blp-trials-short.txt")
```
Glimpse is a part of dplyr: it offers another view of a dataframe.
```{r}
glimpse(blp_df, n=50)
```

# Dplyr verbs
## Select
```{r}
blp_prep <- select(blp_df, participant, lex, resp, rt)
glimpse(blp_prep)
```

other ways of selecting variables:
```{r}
select(blp_df, spell:prev.rt)
select(blp_df, 3:5)
select(blp_df, participant, 4:6)
select(blp_df, starts_with("p"))
select(blp_df, ends_with("ll"))
select(blp_df, contains("sp"))
#if i want "rt" at the beggining of the word - use regex!
select(blp_df, matches("^rt"))
select(blp_df, matches("rt$"))
select(blp_df, matches("^rt|rt$"))
select(blp_df, matches("var[0-9]+$"))#starts with var and ends with numeric value
```
## Deselecting variables
```{r}
select(blp_df, -participant)
select(blp_df, -ends_with("t"))
select(blp_df, -c(2:5))
```
## Reoredering
if we want that the df starts with rt variable.
```{r}
select(blp_df, rt, everything())
```
Select based on condition:
```{r}
select_if(blp_df, is.numeric)
```
```{r}
has_low_mean <- function(x){
  is.numeric(x) && (mean(x, na.rm=T) < 700)
}
```

```{r}
select_if(blp_df, is.numeric)
select_if(blp_df, has_low_mean)
select_if(blp_df, ~is.numeric(.))#using purrr style
```
## Renaming variables:
```{r}
select(blp_df, participant, reaction_time = rt)
```
Otherwise, use `rename` verb for renaming the variables.

```{r}
rename_all(blp_df, ~ str_replace_all(., patter = "\\.", replacement = "_") )#we escape . because regex understands it as any character
```
Renaming only some of the variables, based on the condition, you can use any conditions, also based on character string matches.
```{r}
rename_at(blp_df, 
          vars(matches("^rt|rt$" )),
          ~ str_replace_all(., pattern = "rt", replacement = "reactiontime"))
```
```{r}
rename_at(blp_df, 
          vars(starts_with("p")),
          ~ str_replace_all(., pattern = "rt", replacement = "reactiontime"))

rename_at(blp_df, 
          vars(rt,spell,resp),
          ~ str_replace_all(., pattern = "rt", replacement = "reactiontime"))


rename_if(blp_df, 
          is.character,
          str_to_upper) #here str to upper is a function name not a function call! so
#we dont need ~and (.). 
```
## Slice and filter
Here we will be selecting based on rows, not on the columns.
Slice selects rows by their indicies.
```{r}
c(2,17,37)
slice(blp_df, c(2,17,37))
slice(blp_df, 10:20)
slice(blp_df, seq(10,500,by =  50))
```
if you use negative with slice, you get the inverse (everything other than in the condition)
```{r}
slice(blp_df, -seq(10,500,by =  50))
slice(blp_df, -(5:10)) #this is very useful to remove the first rows, which are often just junk!
slice(blp_df, (nrow(blp_df)-3):nrow(blp_df))#too verbose!
slice(blp_df, (n()-3):n()) #n() - number of rows in the df that is being operated on.
```
### Filter
```{r}
filter(blp_df, lex == "W")# == is the assignment operator in R.
```
Using multiple conditions
```{r}
filter(blp_df, lex == "W", resp == "W", rt < 500)# commas here means #and#
filter(blp_df, lex == "W" & resp == "W"& rt < 500)
filter(blp_df, lex == "W"| resp == "W")#either one or another equal to W.
```
Other examples
```{r}
x <- c(2,3,4,6)
#is nubmer 4 in this set?
4 %in% x#TRUE
```
Return all the rows where spell is any of ...
```{r}
filter(blp_df,rt %in% 500:600)#where rt is between 500 and 600.
```
Complex conditions:
```{r}
filter(blp_df,
lex == 'W',
str_length(spell) < 5 & (resp != lex | rt.raw > 900))
```
Filtering NAs: all the rows that do have NAs.
```{r}
filter_all(blp_df, any_vars(is.na(.)))
#the opposite
filter_all(blp_df, any_vars(!is.na(.)))
```
But if you just get rid of NA, use base R function such as:
na.omit(df)

Filter at:
when selected variables are greater than 500
```{r}
filter_at(blp_df,
          vars(rt:rt.raw),
          all_vars(.> 1000)#the dot is the variables being selected on a particular row.
          )
```
```{r}
filter_at(blp_df,
          vars(rt:rt.raw),
          all_vars(.> median(., na.rm=T))#the dot is the variables being selected on a particular row.
          )
```
```{r}
filter_if(blp_df,
          is.numeric,
          all_vars(. > median(., na.rm=T)))
```
# Mutate
Creating new variables or modifying the old ones
```{r}
mutate(blp_df,
       accurate = lex == resp)#accurate is a logical vector
```
```{r}
mutate(blp_df,
       accurate = lex == resp,
       string_length = str_length(spell))
```
Rescaling the variables with mutate
Here, all variables would be convert to character/as.factor.
```{r}
mutate_all(blp_df,
           as.character)
```

NB! vars( ) is used with mutate_at, 
mutate_if you don't need vars() you can jsut as "is.numeric".
```{r}
mutate_at(blp_df,
          vars(rt:rt.raw),
          scale)

#use the following, if scale returns a matrix and not vector.
mutate_at(blp_df,
          vars(rt:rt.raw),
          ~ as.vector(scale(.)))

#this will not work!
mutate_if(blp_df, 
          vars(is.numeric) #mutate_if do not take vars() as argument
          ~as.vector(scale))
```
```{r}
mutate(blp_df,
       is_fast = ifelse(rt.raw < 700,
                        "fast",
                       "slow")
)
```
Using the recode function to do a similar thing
```{r}
mutate(blp_df,
       lex = recode(lex,"W" = "word", "N" = "noword")) #old value = new value, not like in rename!
```

mutate with case_when - be careful to include all the variables!
```{r}
mutate(blp_df,
prev.rt = case_when(
  prev.rt < 500 ~ 'fast', 
  prev.rt > 1500 ~ 'slow', 
  TRUE ~ 'medium'
) )
```
# Arrange verb - for sorting
```{r}
arrange(blp_df,
        participant,
        rt.raw)
```
```{r}
arrange(blp_df,
        desc(rt.raw))
```
# Summarize verbs
Summarize with summary statistics,
summarize will return a tibble, and tibble outputs a truncated value without decimals. If you do need to see the decimals, change the options.
This is a session-specific setting
```{r}
options(pillar.sigfic = 3)
```

```{r}
summarize(blp_df,
          mean(rt.raw, na.rm = T))
```

to give variable a name:
```{r}
summarize(blp_df,
         avg = mean(rt.raw, na.rm = T),
         median = median(rt.raw, na.rm = T))
```

Scoped version of summarize - if , at, all
Return unique values for each variable
```{r}
summarize_all(blp_df, n_distinct)
```

```{r}
summarise_at(
  blp_df,
  vars(rt:rt.raw),
  list(
    ~ median(., na.rm = T),
    ~ mean(., na.rm = T),
    ~ sd(., na.rm = T)
  )
)
```

Summarize_if summarizes based on TRUE -FALSE of a condition
```{r}
summarize_if(
  blp_df,
  is.numeric,
  list(
    ~ median(., na.rm = T),
    ~ mean(., na.rm = T),
    ~ sd(., na.rm = T)
  )
)
```
# Group_by and pipes
Grouping by participants, as an example.
Looks like the original dataframe, but there are now 78 groups, implicitly.
```{r}
group_by(blp_df, participant)
```
```{r}
summarize(group_by(blp_df,participant),
         avg = mean(rt.raw, na.rm = T),
         median = median(rt.raw, na.rm = T))
```

```{r}
group_by(blp_df,participant, lex) %>% 
summarize(
         avg = mean(rt.raw, na.rm = T),
         median = median(rt.raw, na.rm = T),
         stdev = sd(rt.raw, na.rm = T))
```
```{r}
blp_df %>%
  mutate(
    accuracy = lex == resp,
    string_length = str_length(spell)
  ) %>%
  select(string_length, accuracy, starts_with("rt")
  ) %>% 
  group_by(string_length) %>% 
  summarise(avg_rt = mean(rt.raw, na.rm = T),
          avg_accuracy = mean(accuracy, na.rm = T))
 
```
# Combining data frames

```{r}
Df_1 <- tibble(x = c(1, 2, 3),
              y = c(2, 7, 1), 
              z = c(0, 2, 7))
Df_2 <- tibble(y = c(5, 7), 
               z = c(6, 7), 
               x = c(1, 2))
Df_3 <- tibble(a = c(5, 6, 1), 
               b = c('a', 'b', 'c'),
                     c = c(T, T, F))
```

```{r}
bind_rows(Df_1, Df_2)
```
```{r}
bind_cols(Df_1,Df_3)
```
# Joining df with dplyr
```{r}
stimuli <- read_csv('../data/blp_stimuli.csv')
```

We want to merge this information with blp_df.
The inner_join goes though both dfs and finds a common variable, here - spell. Only the matching rows are kept by inner_join.
```{r}
inner_join(blp_df, stimuli)
```
```{r}
Df_a <- tibble(x = c(1, 2, 3), 
               y = c('a', 'b', 'c'))

Df_b <- tibble(x = c(2, 3, 4), 
               z = c('d', 'e', 'f'))
```

```{r}
inner_join(Df_a, Df_b)
```

Left_join will keep all the values on the df on the left!
```{r}
left_join(Df_a, Df_b)
```
right_join will preserve all the values of the df on the right.
```{r}
right_join(Df_a, Df_b)
```

full_join does right and left join at the same time:
```{r}
full_join(Df_a, Df_b)
```
When names of the variables are different
```{r}
Df_4 <- tibble(x = c(1, 2, 3),
               y = c(2, 7, 1),
               z = c(0, 2, 7))
Df_5 <- tibble(a = c(1, 1, 7),  
               b = c(2, 3, 7),
               c = c('a', 'b', 'c'))
```

```{r}
inner_join(Df_4, Df_5, by = c("x" = "a", "y" = "b"))
inner_join(Df_4, Df_5, by = c("x" = "a"))
```

